---
title: "The Resilience Paradox: Mapping EU-China Strategic Trade and Financial Fragmentation"
subtitle: "2020–2026"
author: "Alex HAAS, Viliam POHANCENIK, Pieter PEVERELLI"
date: last-modified
format:
  html:
    theme: flatly
    toc: true
    code-fold: true
    self-contained: true
  pdf:
    toc: true
    number-sections: true
editor: source
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| label: setup

# 1. Load Libraries
library(knitr)
library(here)
library(dplyr)
library(readr)
library(ggplot2)
library(zoo)
library(lubridate)
library(strucchange)

# 2. Source the Corporate Style
# We use here() to point to the sibling folder "40_Scripts"
source(here::here("40_Scripts/00_style.R"))

# 3. Load Data Globally
# Trade Data
trade_data <- read_csv(here::here("10_Data/11_Processed/01_data_clean_sitc.csv"), 
                       show_col_types = FALSE) %>%
  mutate(date = as.Date(date))

# Finance Data
finance_data <- read_csv(here::here("10_Data/11_Processed/cleaned_BIS_monthly_all_indicators.csv"), 
                         show_col_types = FALSE) %>%
  mutate(date = as.Date(date))
```

## Introduction: The Geoeconomics of De-risking

### From Interdependence to Economic Resilience

The evolution of EU-China economic relations has undergone a fundamental departure from the *Wandel durch Handel* (change through trade) paradigm toward a strategy of "de-risking". This shift is primarily aimed at correcting a trade deficit that reached €304.5 billion by 2024 and mitigating the risks of "coercive leverage" in critical supply chains.

Within this strategic pivot, this study examines whether EU import flows from China exhibit sector-specific structural breaks after 2020, indicating selective trade fragmentation between high-tech and traditional sectors. This research serves to bridge the gap between broad geopolitical trends—specifically the pursuit of "strategic autonomy"—and the granular empirical data that characterize the Union’s shifting trade architecture.

### Addressing Asymmetric Dependencies

Despite this clear political shift toward strategic autonomy, there remains a gap in the empirical literature regarding whether these policy objectives have successfully translated into a measurable divergence in trade flows. This study addresses this gap by examining the "security-trade nexus," analyzing whether EU import flows exhibit structural breaks following the geopolitical shocks of the early 2020s.

We hypothesize a dual-track geoeconomic reality: while traditional, low-value-added sectors governed by market-driven cost efficiencies should remain stable, high-tech industries—subject to the EU Chips Act, foreign investment screenings, and export controls—are expected to show significant fragmentation. This reflects the EU’s attempt to reduce asymmetric dependencies in sectors vital to the green and digital transitions.

```{r}
#| label: fig-trends
#| fig-cap: "3-Month Rolling Average of EU Trade with China by Sector Group. The red dashed line marks the start of the Economic Security Strategy."

# Logic from 05_first_look.R
plot_data <- trade_data %>%
  filter(partner == "CN_X_HK") %>%
  group_by(date, sector_group) %>%
  summarise(total_trade_value = sum(values, na.rm = TRUE), .groups = 'drop') %>%
  group_by(sector_group) %>%
  mutate(
    rolling_avg = rollmean(total_trade_value, k = 6, fill = NA, align = "center")
  )

ggplot(plot_data, aes(x = date, y = rolling_avg, color = sector_group)) +
  geom_line(linewidth = 1.2) + 
  theme_esc() +
  labs(title = "EU Imports from China: Sector Divergence",
       subtitle = "High-Tech vs. Traditional Trade Flows (Rolling Average)",
       x = "Date",
       y = "Trade Value (USD)",
       color = "Sector Group") +
  scale_color_manual(values = esc_colors) +
  geom_vline(xintercept = as.Date("2023-01-01"), 
             linetype = "dashed", 
             color = "#D9534F", 
             linewidth = 1) +
  annotate("text", 
           x = as.Date("2022-12-01"), 
           y = max(plot_data$rolling_avg, na.rm = TRUE),
           label = "Start of Divergence", 
           hjust = 1,
           color = "#D9534F", 
           fontface = "bold",
           size = 4) +
  scale_y_continuous(labels = function(x) paste(x / 1000, "B"))
  
```

## Methodology: Measuring Geoeconomic Re-securitization

### Data Acquisition and Strategic Categorization

The empirical foundation of this research rests on a high-frequency longitudinal dataset (2018–2025) constructed through automated API retrieval from the Eurostat (COMEXT) and Bank for International Settlements (BIS) databases. To test the hypothesis of "selective fragmentation," trade flows are disaggregated using the Standard International Trade Classification (SITC) Revision 4:

-   **Treatment Group (High-Tech & Strategic):** Comprising SITC 5 (Chemicals) and SITC 7 (Machinery/Transport). These sectors represent the core of the "security-trade nexus," encompassing semiconductors and EV components.

-   **Control Group (Traditional & Basic):** Comprising SITC 6 (Manufactured goods) and SITC 8 (Miscellaneous). These sectors serve as a baseline for market-driven trade.

### Econometric Framework: The Chow Test

The primary analytical instrument utilized is the **Chow Test**, a standard in econometric literature for identifying "structural breaks" in time-series data. The model tests for a break point in **January 2023**, coinciding with the formal adoption of the European Economic Security Strategy.

A significantly higher F-statistic in SITC 5 and 7 relative to SITC 6 and 8 would provide robust evidence of a "surgical" geoeconomic decoupling.

```{r}
#| label: analysis-chow
#| results: hide

# Logic from 07_strucchange.R
# Filter for High-Tech Imports
ts_data <- trade_data %>%
  filter(grepl("CN_X_HK", partner)) %>%
  filter(sector_group == "High-Tech & Strategic") %>%
  group_by(date) %>%
  summarise(values = sum(values, na.rm = TRUE)) %>%
  arrange(date)

# Create TS object
ts_val <- ts(ts_data$values, start = c(2020, 1), frequency = 12)

# Run Chow Test at Jan 2023 (2023.0)
break_index <- which(time(ts_val) == 2023.0) 
chow_test <- sctest(ts_val ~ 1, type = "Chow", point = break_index)

# Store results for inline text
f_stat_hightech <- round(chow_test$statistic, 2)
p_val_hightech <- format.pval(chow_test$p.value, eps = .001)

# Logic from 08_strucchange_control.R (Traditional)
ts_data_control <- trade_data %>%
  filter(grepl("CN_X_HK", partner)) %>%
  filter(sector_group == "Traditional & Basic") %>%
  group_by(date) %>%
  summarise(values = sum(values, na.rm = TRUE)) %>%
  arrange(date)

ts_val_ctrl <- ts(ts_data_control$values, start = c(2020, 1), frequency = 12)
chow_test_ctrl <- sctest(ts_val_ctrl ~ 1, type = "Chow", point = break_index)
f_stat_ctrl <- round(chow_test_ctrl$statistic, 2)
# Calculate Intensity Ratio dynamically
intensity_ratio <- round(chow_test$statistic / chow_test_ctrl$statistic, 1)
```

## Discussion and Conclusion: Evidence of Selective Fragmentation

### Empirical Validation of the "Surgical" De-risking Hypothesis

The results of the structural break analysis provide compelling evidence that the European Union’s trade architecture is undergoing a fundamental re-orientation.

For the **Treatment Group (High-Tech & Strategic)**, the model yielded an F-statistic of **`r f_stat_hightech`** (p \< 0.0001). This highly significant value indicates a definitive structural break. Conversely, while the **Control Group** also showed a significant break, the magnitude was markedly lower (F = **`r f_stat_ctrl`**).

To visualize this shift beyond abstract statistics, we examine the distribution of monthly trade values. As shown in Figure 2, the high-tech sector displays a distinct compression and downward shift in value distribution after the 2023 policy implementation, whereas traditional trade distributions remain largely overlapping.

```{r}
#| label: fig-boxplot
#| fig-cap: "Distribution of Monthly Trade Values: Before vs. After the Economic Security Strategy (Jan 2023)."

# 1. Prepare Data with Time Periods
boxplot_data <- trade_data %>%
  filter(partner == "CN_X_HK") %>%
  mutate(
    Period = ifelse(date < as.Date("2023-01-01"), "Pre-Strategy (2020–2022)", "Post-Strategy (2023–)"),
    # Reorder factor so Pre is first
    Period = factor(Period, levels = c("Pre-Strategy (2020–2022)", "Post-Strategy (2023–)"))
  )

# 2. Plot
ggplot(boxplot_data, aes(x = sector_group, y = values, fill = Period)) +
  geom_boxplot(outlier.shape = 21, outlier.fill = "white", alpha = 0.8) +
  scale_fill_manual(values = c("Pre-Strategy (2020–2022)" = "#94a3b8", 
                               "Post-Strategy (2023–)" = "#005f73")) +
  scale_y_continuous(labels = function(x) paste0(round(x / 1000, 1), "B")) +
  labs(
    title = "Structural Shift in Trade Distributions",
    subtitle = "High-Tech trade shows a lower median and compressed range post-2023.",
    x = "",
    y = "Monthly Trade Value (USD)"
  ) +
  theme_esc() +
  theme(legend.position = "top")
```

The Intensity Ratio of approximately `r intensity_ratio` x is the most salient finding of this study. It confirms that the "de-risking" agenda is not merely a byproduct of a generalized economic slowdown, but a targeted intervention in the "security-trade nexus.

### From Market Logic to Political Authority

Recognizing that geoeconomics is inherently multi-disciplinary, we integrate trade data with financial flows from the BIS to capture the "finance-trade nexus".

```{r}
#| label: fig-scatter
#| fig-cap: "Correlation between Financial Exposure and Strategic Trade. The positive slope suggests banks and importers are de-risking in tandem."

# 1. Aggregate High-Tech Trade per month
trade_hightech <- trade_data %>%
  filter(grepl("CN_X_HK", partner), sector_group == "High-Tech & Strategic") %>%
  group_by(date) %>%
  summarise(trade_val = sum(values, na.rm = TRUE))

# 2. Join with Finance Data
scatter_data <- inner_join(trade_hightech, finance_data, by = "date") %>%
  rename(finance_val = values)

# 3. Plot
ggplot(scatter_data, aes(x = finance_val, y = trade_val)) +
  # Add points
  geom_point(color = "#005f73", alpha = 0.6, size = 3) +
  # Add a trend line (Linear Regression)
  geom_smooth(method = "lm", color = "#b91c1c", linetype = "dashed", se = FALSE) +
  # Scales
  scale_x_continuous(labels = function(x) paste0(round(x / 1000, 1), "B")) +
  scale_y_continuous(labels = function(x) paste0(round(x / 1000, 1), "B")) +
  # Labels
  labs(
    title = "The Finance-Trade Nexus",
    subtitle = "Correlation between Eurozone Banking Claims and High-Tech Imports",
    x = "Banking Claims (USD)",
    y = "High-Tech Imports (USD)"
  ) +
  theme_esc()
```

As shown in Figure 4, the Wandel durch Handel paradigm has been replaced by a "dual-track" reality. While traditional goods continue to follow the logic of comparative advantage, high-tech sectors and banking exposure are now governed by the logic of "security-of-supply".


```{r}
#| label: fig-unified
#| fig-cap: "The Dual De-Risking: Trade & Finance Divergence. Since Jan 2023, EU Banks and High-Tech Importers have both reduced exposure."

# Logic from 06_finance_x_imports_CN.R

# 1. Process Trade (Index)
trade_indexed <- trade_data %>%
  filter(grepl("CN_X_HK", partner)) %>%
  filter(sector_group %in% c("High-Tech & Strategic", "Traditional & Basic")) %>%
  group_by(date, sector_group) %>%
  summarise(values = sum(values, na.rm = TRUE), .groups = "drop") %>%
  group_by(sector_group) %>%
  mutate(smooth_value = rollmean(values, k = 3, fill = NA, align = "center")) %>%  
  group_by(sector_group) %>%
  mutate(
    base_val = mean(smooth_value[format(date, "%Y") == "2022"], na.rm = TRUE),
    index_val = (smooth_value / base_val) * 100
  ) %>%
  select(date, sector_group, index_val)

# 2. Process Finance (Index)
finance_indexed <- finance_data %>%
  mutate(
    base_val = mean(values[format(date, "%Y") == "2022"], na.rm = TRUE),
    index_val = (values / base_val) * 100
  ) %>%
  select(date, sector_group, index_val)

# 3. Combine
plot_data_unified <- bind_rows(trade_indexed, finance_indexed)

# 4. Plot
ggplot(plot_data_unified, aes(x = date, y = index_val, color = sector_group, linetype = sector_group)) +
  geom_hline(yintercept = 100, color = "black", linetype = "dotted") +
  geom_vline(xintercept = as.Date("2023-01-01"), color = "#D9534F", linetype = "dashed") +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = esc_colors) +
  scale_linetype_manual(values = c(
    "High-Tech & Strategic" = "solid",
    "Traditional & Basic" = "solid",
    "Financial Exposure (BIS)" = "longdash" 
  )) +
  labs(
    title = "The Dual De-Risking: Trade & Finance Divergence",
    y = "Index (2022 Avg = 100)",
    x = "",
    caption = "Sources: Eurostat (Trade), BIS Locational Banking Statistics (Finance)"
  ) +
  theme_esc()

```


### Forecast and Policy Implications

Looking ahead, we project the "New Normal" trajectory for strategic trade. Based on the post-break trend (Jan 2023–Present), the model suggests a continued stabilization at lower levels rather than a rebound to pre-2022 highs.

```{r}
#| label: fig-forecast
#| fig-cap: "Projecting the 'De-risking' Trend into 2026 based on linear extrapolation of the post-break trajectory."

# Logic from 13_prediction.R
df_hist <- trade_data %>%
  filter(grepl("CN_X_HK", partner)) %>%
  filter(sector_group == "High-Tech & Strategic") %>%
  group_by(date) %>%
  summarise(values = sum(values, na.rm = TRUE)) %>%
  arrange(date)

df_train <- df_hist %>% filter(date >= as.Date("2023-01-01"))
model_lm <- lm(values ~ date, data = df_train)

future_dates <- seq(max(df_hist$date), by = "month", length.out = 13)[-1]
df_future <- data.frame(date = future_dates)
pred <- predict(model_lm, newdata = df_future, interval = "confidence", level = 0.95)

df_forecast_raw <- cbind(df_future, pred) %>%
  rename(values = fit, lower = lwr, upper = upr) %>%
  mutate(type = "Forecast")

# Connect the dots
last_hist_point <- df_hist %>% slice_tail(n = 1) %>% mutate(type = "Forecast", lower = values, upper = values)
df_forecast <- bind_rows(last_hist_point, df_forecast_raw)
df_hist <- df_hist %>% mutate(lower = NA, upper = NA, type = "History")

ggplot() +
  geom_line(data = df_hist, aes(x = date, y = values), color = "#005f73", linewidth = 1.2) +
  geom_line(data = df_forecast, aes(x = date, y = values), color = "#005f73", linetype = "dashed", linewidth = 1.2) +
  geom_ribbon(data = df_forecast, aes(x = date, ymin = lower, ymax = upper), fill = "#005f73", alpha = 0.15) +
  geom_vline(xintercept = as.Date("2023-01-01"), linetype = "dashed", color = "#D9534F", linewidth = 1.2) +
  scale_y_continuous(labels = function(x) paste0(round(x / 1000, 1), " B")) + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Forecast 2026: Strategic Sector Trajectory", x = "Date", y = "Trade Value (USD)") +
  theme_esc()
```

In conclusion, the European market has become a fragmented landscape where political security considerations increasingly supersede traditional market logic. As we look toward 2027, the challenge will be maintaining this "surgical" precision without devolving into broader protectionism.

# References

1.  Bank for International Settlements. (2026). Locational banking statistics.

2.  Chow, G. C. (1960). Tests of equality between sets of coefficients in two linear regressions. *Econometrica*.

3.  European Commission. (n.d.). China Trade Relationships.

4.  Eurostat. (2026). Euro area trade by SITC product group.

5.  McNamara, K. R. (2024). Transforming Europe? The EU’s industrial policy and geopolitical turn. *Journal of European Public Policy*.

6.  Pereira, I. T. (2025). What are the most imported and exported products between China and the EU? *Euronews*.

7.  Vandermeeren, F. (2024). Understanding EU-China economic exposure. *European Commission*.
